---
title: "Wikidata Ontology vs Wikipedia Categories"
author: "Emma Leonhart, Evan Pasenau, Jordyn Campen"
format: html
---

## Introduction

This report compares Wikidata's P31 (instance of) ontological classification with English Wikipedia's category system across multiple domains. The pipeline runs periodically via GitHub Actions, re-querying both APIs to capture changes over time.

## Data Acquisition

```{python}
import pandas as pd
import os
from datetime import datetime

df = pd.read_csv("../data/processed/ontology_comparison.csv")
summary = pd.read_csv("../data/processed/domain_summary.csv")

# Show when the data file was last modified
mtime = os.path.getmtime("../data/processed/ontology_comparison.csv")
print(f"Data snapshot: {datetime.fromtimestamp(mtime).strftime('%Y-%m-%d %H:%M UTC')}")
print(f"Total items: {len(df)}")
print(f"Domains: {df['domain'].nunique()}")
```

## Domain Summary

```{r}
library(ggplot2)

summary <- read.csv("../data/processed/domain_summary.csv")

ggplot(summary, aes(x = reorder(domain, -mean_overlap_ratio), y = mean_overlap_ratio)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Mean P31-Category Overlap by Domain",
    x = "Domain",
    y = "Mean Overlap Ratio"
  ) +
  theme_minimal()
```

## P31 Count vs Category Count

```{r}
df <- read.csv("../data/processed/ontology_comparison.csv")

ggplot(df, aes(x = p31_count, y = category_count, color = domain)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "P31 Count vs Wikipedia Category Count",
    x = "Number of P31 (instance of) values",
    y = "Number of Wikipedia categories"
  ) +
  theme_minimal()
```

## Overlap Distribution

```{r}
ggplot(df, aes(x = domain, y = overlap_ratio, fill = domain)) +
  geom_boxplot() +
  labs(
    title = "Overlap Ratio Distribution by Domain",
    x = "Domain",
    y = "Overlap Ratio (P31 labels found in categories)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Findings

_TODO: Interpret the results after data acquisition._

## Ethics and Limitations

- **Cultural bias**: Only English Wikipedia is analyzed, introducing Anglophone cultural bias.
- **API rate limits**: Exponential backoff and local caching prevent IP bans.
- **Snapshot in time**: Both Wikidata and Wikipedia change constantly; results reflect the data at acquisition time.
